Interval-Valued Quantum Probabilities: Contextuality and the Born Rule

### 2017-05-15 prop_letter.tex
small fixes

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain events~\cite{Kolmogorov1950}.  In the quantum case, the
events of interest are the \emph{projection operators} $P$. This set
includes the empty projection~$\mathbb{0}$, the identity
projection~$\mathbb{1}$, projections of the form $\proj{\psi}$ where
$\ket{\psi}$ is a pure quantum state, sums of \emph{orthogonal}
projections, and products of \emph{commuting} projections. In a
well-defined quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event $P_i$ is associated with a probability $\mu(P_i)$ subject
to the following axioms: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for a set of
mutually orthogonal projections $\left\{ P_{i}\right\} _{i=1}^{N}$, we
have:
\begin{equation}
\mu\left(\sum_{i=1}^{N}P_{i}\right)=
\sum_{i=1}^{N}\mu\left(P_{i}\right)
\textrm{ .}\label{eq:QuantumProbability-Addition}
\end{equation}

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable $Q$
with spectral decomposition $\proj{0} + 2\,\proj{1} +
3\,\proj{2}$. Here are two fragments of valid probability measures
$\mu_1$ and $\mu_2$ that can associated with this space:
$\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule,
the first probability measure corresponds to the quantum system being
in the state $(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the second corresponds
to the quantum system being in the state
$(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where $|\alpha|=1$. The expectation value
of the observable in the first case is 1.5 and in the second it is 2.

@@

To adapt the axioms of such IVPMs to the quantum case, we proceed as
follows. First we fix a collection of intervals $\mathscr{I}$ that
represents an abstraction of the available experimental resources. At
one extreme, infinitely precise resources would correspond to the
collection of intervals $\{[x,x] \mid x \in [0,1]\}$. This case should
coincide with the idealized situation above. At another extreme, an
experiment with \emph{no} resources would correspond to the following
three intervals $\imposs=\left[0,0\right]$,
$\necess=\left[1,1\right]$, and $\unknown=[0,1]$. As before, the first
two intervals are associated with the empty projection and the
identity projection but every other projection would be associated
with $\unknown$; its probability can be anything in the range
$[0,1]$. A third special situation would be to consider IVPMs with
only the two intervals $\imposs=\left[0,0\right]$ and
$\necess=\left[1,1\right]$. This is what Peres calls the
\emph{crypto-deterministic} case in which there is no ambiguity about
the status of any event; we know for certain if it will occur or not.

Given a collection of intervals $\mathscr{I}$ with addition and scalar
multiplication defined as follows: 
\begin{subequations}\label{eq:interval-operations}
\begin{eqnarray}
 &  & [l_{1},r_{1}]+[l_{2},r_{2}]=[l_{1}+l_{2},r_{1}+r_{2}]\textrm{ and}\\
 &  & x[l,r]=\begin{cases}
[xl,xr] & \textrm{for }x\ge0\textrm{ ;}\\{}
[xr,xl] & \textrm{for }x\le0\textrm{ ,}
\end{cases}
\end{eqnarray}
\end{subequations}
a QIVPM $\bar{\mu}$ is an assignment of intervals to each event
obeying the following axioms:
$\bar{\mu}(\mathbb{0})=\imposs=\left[0,0\right]$,
$\bar{\mu}(\mathbb{1})=\necess=\left[1,1\right]$,
$\bar{\mu}\left(\mathbb{1}-P\right)=\left[1,1\right]-\bar{\mu}\left(P\right)$,
and for a set of mutually orthogonal projections
$\left\{ P_{i}\right\} _{i=1}^{N}$, we have:
\begin{equation}
\bar{\mu}\left(\sum_{i=1}^{N}P_{i}\right)\subseteq\sum_{i=1}^{N}\bar{\mu}\left(P_{i}\right)\label{eq:QuantumInterval-valuedProbability-Inclusion}
\end{equation}
The first three axioms are the direct counterpart of the corresponding
axioms for conventional probability measures. The last condition
allows for the interval associated with a composite event to be
sharper than the interval calculated from each individual
event. Intuitively, consider an event $P$ and its complement
$\mathbb{1}-P$. Looking at each separately we may discover
$\bar{\mu}(P) = [0.2,0.3]$ and $\bar{\mu}(\mathbb{1}-P)=[0.7,0.8]$. By
adding the individual intervals, we predict the probability of the
composite event $P+(\mathbb{1}-P)$ to be in the range
$[0.9,1.1]$. Eq.~\ref{eq:QuantumInterval-valuedProbability-Inclusion}
allows us to refine this to $[1,1]$ by recognizing that the composite
event is the identity projection.

@@

@@

Yu-Tsung: please construct an example with commuting events P0 and P1 such that:

@@

and $[x,y]$ is NOT included in $[z,w]$

>>>>

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain events~\cite{Kolmogorov1950}.  In the quantum case, the
events of interest are specified by \emph{projection operators}
$P$. This set $P$ includes the empty projection~$\mathbb{0}$, the
identity projection~$\mathbb{1}$, projections of the form
$\proj{\psi}$ where $\ket{\psi}$ is a pure quantum state, sums of
\emph{orthogonal} projections, and products of \emph{commuting}
projections. In a well-defined quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event $P_i$ is associated with a probability $\mu(P_i)$ subject
to the following axioms: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for a set of
mutually orthogonal projections $\left\{ P_{i}\right\} _{i=1}^{N}$, we
have:
\begin{equation}
\mu\left(\sum_{i=1}^{N}P_{i}\right)=
\sum_{i=1}^{N}\mu\left(P_{i}\right)
\textrm{ .}\label{eq:QuantumProbability-Addition}
\end{equation}

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable $Q$
with spectral decomposition $\proj{0} + 2\,\proj{1} +
3\,\proj{2}$. Two fragments of valid probability measures $\mu_1$ and
$\mu_2$ that can associated with this space: $\mu_1(\proj{0}) = 0.5$,
$\mu_1(\proj{1}) = 0.5$, $\mu_1(\proj{2}) = 0$, and
$\mu_2(\proj{0}) = 0.5$, $\mu_2(\proj{1}) = 0$,
$\mu_2(\proj{2}) = 0.5$. By the Born rule, the first probability
measure corresponds to the quantum system being in the state
$(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the second corresponds to the
quantum system being in the state $(\ket{0}+\alpha\ket{2})/\sqrt{2}$,
where $|\alpha|=1$. The expectation value of the observable in the
first case is 1.5 and in the second it is 2.

@@

To adapt the axioms of such IVPMs to the quantum case, we proceed as
follows. First we fix a collection of intervals $\mathscr{I}$ that
represents an abstraction of the available experimental resources. At
one extreme, infinitely precise resources would correspond to the
collection of intervals $\{[x,x] \mid x \in [0,1]\}$. This case should
coincide with the idealized situation above. At another extreme, an
experiment with ``minimal'' resources would correspond to the
following three intervals $\imposs=\left[0,0\right]$,
$\necess=\left[1,1\right]$, and $\unknown=[0,1]$. As before, the first
two intervals are associated with the empty projection and the
identity projection but every other projection would be associated
with $\unknown$; its probability can be anything in the range
$[0,1]$. A third special situation would be to consider IVPMs with
only the two intervals $\imposs=\left[0,0\right]$ and
$\necess=\left[1,1\right]$. This is what Peres calls the
\emph{crypto-deterministic} case in which there is no ambiguity about
the status of any event; we know for certain if it will occur or not.

Given a collection of intervals $\mathscr{I}$ with addition and scalar
multiplication defined as follows: 
\begin{subequations}\label{eq:interval-operations}
\begin{eqnarray}
 &  & [l_{1},r_{1}]+[l_{2},r_{2}]=[l_{1}+l_{2},r_{1}+r_{2}]\textrm{ and}\\
 &  & x[l,r]=\begin{cases}
[xl,xr] & \textrm{for }x\ge0\textrm{ ;}\\{}
[xr,xl] & \textrm{for }x\le0\textrm{ ,}
\end{cases}
\end{eqnarray}
\end{subequations}
a QIVPM $\bar{\mu}$ is an assignment of intervals to each event
obeying the following axioms:
$\bar{\mu}(\mathbb{0})=\imposs=\left[0,0\right]$,
$\bar{\mu}(\mathbb{1})=\necess=\left[1,1\right]$,
$\bar{\mu}\left(\mathbb{1}-P\right)=\left[1,1\right]-\bar{\mu}\left(P\right)$,
and for a set of mutually orthogonal projections
$\left\{ P_{i}\right\} _{i=1}^{N}$, we have:
\begin{equation}
\bar{\mu}\left(\sum_{i=1}^{N}P_{i}\right)\subseteq\sum_{i=1}^{N}\bar{\mu}\left(P_{i}\right)\label{eq:QuantumInterval-valuedProbability-Inclusion}
\end{equation}
The first three axioms are the direct counterpart of the corresponding
axioms for conventional probability measures. The last condition
allows for the interval associated with a composite event to be
sharper than the interval calculated from each individual
event. Intuitively, consider an event $P$ and its complement
$\mathbb{1}-P$. Looking at each separately we may discover as a result
of an experiment that $\bar{\mu}(P) = [0.2,0.3]$ and
$\bar{\mu}(\mathbb{1}-P)=[0.7,0.8]$. By adding the individual
intervals, we predict the probability of the composite event
$P+(\mathbb{1}-P)$ to be in the range
$[0.9,1.1]$. Eq.~\ref{eq:QuantumInterval-valuedProbability-Inclusion}
allows us to refine this to $[1,1]$ by recognizing that the composite
event is the identity projection.

@@

\amr{does the following rely on convexity?}

@@

\amr{
Yu-Tsung: please construct an example with commuting events P0 and P1 such that:

@@

and $[x,y]$ is NOT included in $[z,w]$
}
